---
title: "Predicting Goal-to-go Offensive Play Call Probability"
author: "Jake Slaton"
output:
        html_document:
                toc: true
                toc_float: true
                toc_depth: 4
                number_sections: false
---

__Scenario:__ the team is heading into Week 18 and making a push for the playoffs. The defensive coordinator is looking to fortify his play calling in goal-to-go situations. He requests some analyses on the probability of a given play being either a run or a pass, so that he may be able to best prepare his defense in practice and fine-tune his game plan.


```{r setup}
knitr::opts_chunk$set(echo = TRUE,
                      message = FALSE)

## Load necessary packages
library(tidyverse)
library(janitor)
library(ggstats)
library(gtsummary)
library(gt)
library(nflfastR)
library(rstanarm)
library(bayesplot)
library(tidybayes)
library(broom.mixed)
```


```{r data-wrangling}
## 2022 NFL play-by-play data
pbp_raw <- load_pbp(2024)

## Data frame for goal-to-go data
goal_offense <- pbp_raw %>% 
    filter(
        !is.na(goal_to_go),
        !is.na(yardline_100),
        !is.na(play_type),
        goal_to_go == 1,
        play_type %in% c("pass", "run")
    ) %>% 
    mutate(
        down = as.factor(down),
        play_type = factor(play_type, levels = c("run", "pass"))
    ) %>% 
    select(
        posteam,
        week,
        goal_to_go,
        yardline_100,
        down,
        play_type,
        defteam,
        week
    )


```

# Analysis

## EDA 

After querying the necessary data, a good start would be to perform an exploratory data analysis.

```{r league-eda}
## Calculate play call proportions conditional on the down
goal_offense %>% 
    tbl_cross(
        row = down,
        col = play_type,
        label = c(play_type ~ "Play type",
                   down ~ "Down"),
        percent = "row",
        margin = "row"
    ) %>% 
    as_gt() %>% 
    tab_header(title = "2024 Season", subtitle = "Weeks 1 & 2")

## Bar plot to visualize play call proportions for league
goal_offense %>% 
    tabyl(down, play_type) %>% 
    pivot_longer(
        cols = -down,
        names_to = "play_type",
        values_to = "count") %>% 
    ggplot(
        aes(x = down, fill = play_type, weight = count, by = down)
    ) +
    geom_bar(position = "fill") +
    geom_text(position = position_fill(vjust = 0.5), stat = "prop") +
    scale_y_continuous(labels = scales::percent) +
    scale_fill_manual(
        values = c("skyblue2", "salmon2"),
        labels = c("Pass", "Run")
    ) +
    labs(
        title = "2024 Play Call Proportions By Down",
        x = "Down",
        y = "",
        fill = "Play Call"
    ) +
    theme_minimal()

        
```

As may be assumed, during the 2024 season the proportion of pass plays called in goal-to-go situations increases as the down increases. It may also be important to consider what yard line the line of scrimmage is at, expressed as yards from the end zone.

```{r league-glm-plot}
## Create a scatter plot with a logit curve to display pass probability vs. LOS
goal_offense %>% 
    ggplot(aes(x = yardline_100, y = as.numeric(play_type)-1)) +
    geom_point(
        color = "blue4",
        position = position_jitter(width = 0.5, height = 0.015),
        alpha = 0.3,
        size = 2
    ) +
    geom_smooth(
        method = "glm",
        se = FALSE,
        method.args = list(family = "binomial"),
        color = "salmon2",
        fill = "salmon4",
        linewidth = 2
    ) +
    labs(
        title = "Probability of a Pass Given LOS",
        x = "Yardline",
        y = "Pass Probability"
    ) +
    theme_minimal()
```

As may be expected, the probability of the play being a pass increases as the line of scrimmage becomes further from the goal-line. For example, the expected probability of pass with goal-to-go on the 5 yard line is  ~50%, whereas for the 10 yard line the expected pass probability is ~ 65%. (Since we have play time as binary, this could be ~50% and ~35% run probability for the 5 and 10 yard lines, respectively). Something to consider, however, is how this may be conditional on the down.

```{r league-glm-plot-downs}
## Create a scatter plot with a logit curve to display pass
## probability vs. LOS conditioned on down
goal_offense %>% 
    ggplot(aes(x = yardline_100, y = as.numeric(play_type)-1)) +
    geom_point(
        color = "blue4",
        position = position_jitter(width = 0.5, height = 0.015),
        alpha = 0.3,
        size = 2
    ) +
    geom_smooth(
        aes(color = down),
        method = "glm",
        se = FALSE,
        method.args = list(family = "binomial"),
        linewidth = 2
    ) +
    labs(
        title = "Probability of a Pass Given LOS",
        x = "Yardline",
        y = "Pass Probability",
        color = "Down"
    ) +
    theme_minimal()
```

The probability of pass, given the LOS, is conditional on the down. As seen in the plot, if the LOS is at the 5 yard line, the probability of a pass is ~25%, ~50%, ~80%, ~80% for 1st, 2nd, 3rd, and 4th down, respectively.

## 2024 weakly informative

Now that the data has been explored and a direction provided, statistical modeling will come into play. Given the visuals, it seems a model that predicts play call probability will want the down and the yard line as input variables. A logistic regression model should work for this purpose. Will leave the priors weakly informative to "let the data talk."

```{r 2024-model-default-priors}
## Logistic regression model
fit_bayes_default <- stan_glm(
    play_type ~ 1 + down + yardline_100,
    family = binomial,
    data = goal_offense,
    prior_intercept = normal(0, 2.5, autoscale = TRUE),
    prior = normal(0, 2.5, autoscale = TRUE),
    chains = 4,
    iter = 5000*2,
    seed = 1995
)

## Posterior simulations
fit_bayes_default_sims <- as.data.frame(fit_bayes_default)

## Model results
fit_bayes_default_df <- tidy(fit_bayes_default, conf.int = TRUE, conf.level = 0.95)
fit_bayes_default_df[,2:5] <- lapply(
    X = fit_bayes_default_df[,2:5],
    FUN = function(x) round(x, 3)
)

## Model results but prettier
fit_bayes_default_table <- fit_bayes_default_df %>% 
    gt() %>% 
    cols_label(
        term = "Coefficient",
        estimate = "Estimate",
        std.error = "SE",
        conf.low = "95% LCI",
        conf.high = "95% UCI"
    ) %>% 
    tab_header(title = "2024 Model", subtitle = "Using weak priors")

fit_bayes_default_table

```

We can use past data to gather prior information.

```{r 2021-2023-data}
## Load 2021-2023 data
pbp_21_23 <- load_pbp(2021:2023)
goal_offense_21_23 <- pbp_21_23 %>% 
    filter(
        !is.na(goal_to_go),
        !is.na(yardline_100),
        !is.na(play_type),
        goal_to_go == 1,
        play_type %in% c("pass", "run")
    ) %>% 
    mutate(
        down = as.factor(down),
        play_type = factor(play_type, levels = c("run", "pass"))
    ) %>% 
    select(
        posteam,
        week,
        goal_to_go,
        yardline_100,
        down,
        play_type,
        defteam,
        week
    )

## Calculate play call proportions conditional on the down
goal_offense_21_23 %>% 
    tbl_cross(
        row = down,
        col = play_type,
        label = c(play_type ~ "Play type",
                   down ~ "Down"),
        percent = "row",
        margin = "row"
    ) %>% 
    as_gt() %>% 
    tab_header(title = "2021-2023 Seasons")
```

## 2021-2023 Model

Bayesian model with data from 2021-2023. Default priors.

```{r previous-model}
## Previous (2021-2023) Bayesion model with default priors
fit_bayes_21_23 <- stan_glm(
    play_type ~ 1 + down + yardline_100,
    family = binomial,
    data = goal_offense_21_23,
    prior_intercept = normal(0, 2.5, autoscale = TRUE),
    prior = normal(0, 2.5, autoscale = TRUE),
    chains = 4,
    iter = 5000*2,
    seed = 1995
)

## Results
fit_bayes_21_23_df <- tidy(fit_bayes_21_23, conf.int = TRUE, conf.level = 0.95)
fit_bayes_21_23_df[,2:5] <- lapply(
    X = fit_bayes_21_23_df[,2:5],
    FUN = function(x) round(x, 3)
)

## Results but make it nice
fit_bayes_21_23_table <- fit_bayes_21_23_df %>% 
    gt() %>% 
    cols_label(
        term = "Coefficient",
        estimate = "Estimate",
        std.error = "SE",
        conf.low = "95% LCI",
        conf.high = "95% UCI"
    ) %>% 
    tab_header(title = "2021-2023 Model")

fit_bayes_21_23_table
```

## 2024 Informative

Bayesian model with informative priors

```{r 2024-model-inform-priors}
## Logistic regression model with informative priors
fit_bayes_informative <- stan_glm(
    play_type ~ 1 + down + yardline_100,
    family = binomial,
    data = goal_offense,
    prior_intercept = normal(0, 0.05),
    prior = c(normal(0.85, 0.12),
              normal(1.75, 0.15),
              normal(1.70, 0.30),
              normal(0.13, 0.02)),
    chains = 4,
    iter = 5000*2,
    seed = 1995
)

## Posterior simulations
fit_bayes_informative_sims <- as.data.frame(fit_bayes_informative)


## Model results
fit_bayes_informative_df <- tidy(
    fit_bayes_informative,
    conf.int = TRUE,
    conf.level = 0.95)

fit_bayes_informative_df[,2:5] <- lapply(
    X = fit_bayes_informative_df[,2:5],
    FUN = function(x) round(x, 3)
)

## Model results makeover
fit_bayes_informative_table <- fit_bayes_informative_df %>% 
    gt() %>% 
    cols_label(
        term = "Coefficient",
        estimate = "Estimate",
        std.error = "SE",
        conf.low = "95% LCI",
        conf.high = "95% UCI"
    ) %>% 
    tab_header(title = "2024 Model", subtitle = "Using informative priors")

fit_bayes_informative_table
```

Now to compare the posteriors

```{r posterior-viz-comp}
## Master data frame with both posterior simulations
fit_bayes_posterior_comp <- bind_rows(
    fit_bayes_default_sims,
    fit_bayes_informative_sims
) %>% 
    mutate(
        priors = as.factor(rep(c("Weakly informative", "Informative"), each = 20000))
    )

## Visualize posteriors
fit_bayes_posterior_comp_plot <- fit_bayes_posterior_comp %>% 
    select(-`(Intercept)`) %>% 
    pivot_longer(
        cols = -priors,
        names_to = "parameter"
    ) %>% 
    ggplot(aes(x = value)) +
    geom_density(
        aes(color = priors, fill = priors),
        linewidth = 1.5,
        alpha = 0.4
    ) +
    facet_wrap(~parameter, scales = "free") +
    labs(
        title = "Posterior parameter estimations",
        x = "Coefficient value"
    ) +
    theme_minimal() +
    theme(
        strip.background = element_rect(fill = "grey"),
        panel.border = element_rect(color = "black", fill = NA),
        legend.position = "bottom"
    )

fit_bayes_posterior_comp_plot
```

